Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2021-07-04 source/sec_intro.tex
Update on Overleaf.

\section{Introduction}
Federated learning (FL) is a machine learning setting where many clients (e.g. mobile devices or organizations) collaboratively train a model under the orchestration of a central server (e.g. service provider), while keeping the training data decentralized~\cite{smith2017federated, kairouz2019advances}. In recent years, FL has swiftly emerged as an important learning paradigm~\cite{mcmahan2016communication,li2018federated}--one that enjoys widespread success in applications such as personalized recommendation~\cite{chen2018federated}, virtual assistant~\cite{lamautonomy}, and keyboard prediction~\cite{47586}, to name a few--for at least two reasons: First, the rapid proliferation of smart devices that are equipped with both computing power and data-capturing capabilities provided the infrastructure core for FL. Second, the rising awareness of privacy and the exponential growth of computational power (blessed by Moore's law) in mobile devices have made it increasingly attractive to push the computation to the edge.

Despite its promise and broad applicability in our current era, the potential value FL delivers is coupled with the unique challenges it brings forth. In particular, when FL learns a single statistical model using data from across all the devices while keeping each individual device's data isolated (and hence protects privacy)~\cite{kairouz2019advances}, it faces two challenges that are absent in centralized optimization and distributed (stochastic) optimization~\cite{zhou2017convergence, stich2018local,khaled2019first,liang2019variance,wang2018cooperative,woodworth2018graph,wang2019adaptive,jiang2018linear,yu2019parallel,yu2019linear, khaled2020tighter,koloskova2020unified}:

1) \textbf{Data heterogeneity:} data distributions in devices are different (and data can't be shared);

2) \textbf{System heterogeneity:} only a subset of devices may access the central server at each time both because the communications bandwidth profiles vary across devices and because there is no central server that has control over when a device is active.

% Gaps in the theoretical understanding. 
To address these challenges, Federated Averaging (FedAvg)~\cite{mcmahan2016communication} was proposed as a particularly effective heuristic, which has enjoyed great empirical success~\cite{47586}. This success has since motivated a growing line of research efforts into understanding its theoretical convergence guarantees in various settings. For instance, \cite{haddadpour2019convergence} analyzed FedAvg (for non-convex smooth problems satisfying PL conditions) under the assumption that each local device's minimizer is the same as the minimizer of the joint problem (if all devices' data is aggregated together), an overly restrictive assumption.
Very recently, \cite{li2019convergence} furthered the progress and established an $\cO(\frac{1}{T})$ convergence rate for FedAvg for strongly convex smooth problems. 
At the same time, \cite{huo2020faster} studied the Nesterov accelerated FedAvg for non-convex smooth problems and established 
an $\cO(\frac{1}{\sqrt{T}})$ convergence rate to stationary points.
% (Prior to that, \cite{liu2019accelerating} studied Nesterov accelerated FedAvg using full gradient--rather than stochastic gradient as in both the above-mentioned works and our setting--for strongly convex smooth problems and established an $\cO(\frac{1}{T})$ convergence rate).

However, despite these very recent fruitful pioneering efforts into understanding the theoretical convergence properties of FedAvg, it remains open as to how the number of devices--particularly the number of devices that participate in the computation--affect the convergence speed.
In particular, do we get linear speedup of FedAvg? What about when FedAvg is accelerated? These aspects are currently unexplored in FL. We fill in the gaps here by providing affirmative answers.

\begin{table}[t!]
\hspace{-1em}
{\small
\centering
\begin{tabular}{|c|c|c|c|c|}\hline 
	\multirow{2}{*}{\backslashbox{{\tiny Participation} }{{\tiny Objective function}}} & \multirow{2}{*}{Strongly Convex}     &\multirow{2}{*}{Convex}    & Overparameterized & Overparameterized \\ 
	                                &                        &         &     general case                 & linear regression   \\ \hline 
	Full                         & $\cO(\frac{1}{NT}+\frac{E^{2}}{T^{2}})$    &  $\mathcal{O}\left(\frac{1}{\sqrt{NT}}+\frac{NE^{2}}{T}\right)$   & $\cO(\exp(-\frac{NT}{E\kappa_1}))$ & $\cO(\exp(-\frac{NT}{E\kappa }))^{\dagger}$    \\ \hline
	Partial                      &  $\cO\left(\frac{E^{2}}{KT}+\frac{E^{2}}{T^{2}}\right)$   &  $\cO\left(\frac{E^2}{\sqrt{KT}}+\frac{KE^2}{T} \right)$ &  $\cO(\exp(-\frac{KT}{E\kappa_1}))$ & $\cO(\exp(-\frac{KT}{E\kappa }))^{\dagger}$     \\ \hline
\end{tabular}
}
\caption{Our convergence results for FedAvg and accelerated FedAvg in this paper. Throughout the paper, $N$ is the total
number of local devices, and $K \leq N$ is the maximal number of devices that are accessible to the central server. $T$ is the total number of stochastic updates performed by each local device, $E$ is the local steps between two consecutive server communications (and hence $T/E$ is the number of communications). $^{\dagger}$ In the linear regression setting, we have $\kappa=\kappa_1$ for FedAvg and $\kappa=\sqrt{\kappa_1\tilde{\kappa}}$ for accelerated FedAvg, where $\kappa_1$ and $\sqrt{\kappa_1\tilde{\kappa}}$ are condition numbers defined in Section \ref{sec:overparameterized}. Since
$\kappa_{1}\geq\tilde{\kappa}$, this implies a speedup factor of
$\sqrt{\frac{\kappa_{1}}{\tilde{\kappa}}}$ for accelerated FedAvg.}
% {\raggedright 
% $^{*}$ The fourth column presents the results for the general overparaterized setting. The fifth column presents the results for overparaterized linear regression.
        %   \par}
\label{tb:convergencerateintro}
\vspace{-2.5em}
\end{table}

\textbf{Our Contributions}
We provide a comprehensive convergence analysis
of FedAvg and its accelerated variants in the presence of both data and system heterogeneity. 
Our contributions are threefold.

First, we establish an {\small{$\cO(1/KT)$}} convergence rate  under FedAvg for strongly convex and smooth problems and  an
{\small{$\cO(1/\sqrt{KT})$}} convergence rate for convex
and smooth problems (where $K$ is the number of participating devices), thereby establishing that FedAvg enjoys the desirable linear speedup property in the FL setup. Prior to our work here, the best and the most related convergence analysis is given by \cite{li2019convergence}, which established an $\cO(\frac{1}{T})$ convergence rate for strongly convex smooth problems under FedAvg. Our rate matches the same (and optimal) dependence on $T$, but also completes the picture by establishing the linear dependence on $K$.

Second, we establish the same convergence rates--{\small{$\cO(1/KT)$}} for strongly convex and smooth problems and {\small{$\cO(1/\sqrt{KT})$}} for convex and smooth problems--for Nesterov accelerated FedAvg. 
We analyze the accelerated version of FedAvg here because empirically it tends to perform better; yet, its theoretical convergence guarantee is unknown. To the best of our knowledge, these are the first results that provide a linear speedup characterization of Nesterov accelerated FedAvg in those two problem classes (that FedAvg and Nesterov accelerated FedAvg share the same convergence rate is to be expected: this is the case even for centralized stochastic optimization).

Third, we study a subclass of strongly convex smooth problems where the objective is over-parameterized and establish 
a faster $\cO(\exp(-\frac{KT}{\kappa}))$ convergence rate for FedAvg. Within this class, we further consider the linear regression problem and establish an even sharper rate under FedAvg. In addition, we propose a new variant of accelerated FedAvg--MaSS accelerated FedAvg--and establish a faster convergence rate (compared to if no acceleration is used). This stands in contrast to generic (strongly) convex stochastic problems where theoretically no rate improvement is obtained when one accelerates FedAvg.
The detailed convergence results are summarized in Table~\ref{tb:convergencerateintro}.

>>>>

\section{Introduction}
Federated learning (FL) is a machine learning paradigm where many clients (e.g., mobile devices or organizations) collaboratively train a model under the orchestration of a central server (e.g., service provider), while keeping the training data decentralized (\cite{smith2017federated, kairouz2019advances}). In recent years, FL has swiftly emerged as an important learning paradigm (\cite{mcmahan2016communication,li2018federated})--one that enjoys widespread success in applications such as personalized recommendation (\cite{chen2018federated}), virtual assistant (\cite{lam2019protecting}), and keyboard prediction (\cite{47586}), to name a few--for at least three reasons: First, the rapid proliferation of smart devices that are equipped with both computing power and data-capturing capabilities provided the infrastructure core for FL. Second, the rising awareness of privacy and the explosive growth of computational power in mobile devices have made it increasingly attractive to push the computation to the edge. Third, the empirical success of communication-efficient FL algorithms has enabled increasingly larger-scale parallel computing and learning with less communication overhead.

Despite its promise and broad applicability in our current era, the potential value FL delivers is coupled with the unique challenges it brings forth. In particular, when FL learns a single statistical model using data from across all the devices while keeping each individual device's data isolated (\cite{kairouz2019advances}), it faces two challenges that are absent in centralized optimization and distributed (stochastic) optimization (\cite{zhou2017convergence, stich2018local,khaled2019first,liang2019variance,wang2018cooperative,woodworth2018graph,wang2019adaptive,jiang2018linear,yu2019parallel,yu2019linear, khaled2020tighter,koloskova2020unified,woodworth2020local,woodworth2020minibatch}):

1) \textbf{Data (statistical) heterogeneity:} data distributions in devices are different (and data cannot be shared);

2) \textbf{System heterogeneity:} only a subset of devices may access the central server at each time both because the communications bandwidth profiles vary across devices and because there is no central server that has control over when a device is active (the presence of ``stragglers''). 
% Gaps in the theoretical understanding.

To address these challenges, Federated Averaging (FedAvg)~\cite{mcmahan2016communication} was proposed as a particularly effective heuristic, which has enjoyed great empirical success. This success has since motivated a growing line of research efforts into understanding its theoretical convergence guarantees in various settings. For instance, \cite{haddadpour2019convergence} analyzed FedAvg (for non-convex smooth problems satisfying PL conditions) under the assumption that each local device's minimizer is the same as the minimizer of the joint problem (if all devices' data is aggregated together), an overly restrictive assumption that restricts the extent of data heterogeneity.
Very recently, \cite{li2019convergence} furthered the progress and established an $\cO(\frac{1}{T})$ convergence rate for FedAvg for strongly convex smooth Federated learning problems with both data and system heterogeneity. A similar result in the same setting~\cite{karimireddy2019scaffold} also established an $\cO(\frac{1}{T})$ result that allows for a linear speedup when the number of participating devices is large.
At the same time, \cite{huo2020faster} studied the Nesterov accelerated FedAvg for non-convex smooth problems and established 
an $\cO(\frac{1}{\sqrt{T}})$ convergence rate to stationary points. 
% (Prior to that, \cite{liu2019accelerating} studied Nesterov accelerated FedAvg using full gradient--rather than stochastic gradient as in both the above-mentioned works and our setting--for strongly convex smooth problems and established an $\cO(\frac{1}{T})$ convergence rate).

However, despite these very recent fruitful pioneering efforts into understanding the theoretical convergence properties of FedAvg, it remains open as to how the number of devices--particularly the number of devices that participate in the computation--affects the convergence speed.
In particular, is linear speedup of FedAvg a universal phenomenon across different settings and for any number of devices? What about when FedAvg is accelerated with momentum updates? Does the presence of both data and system heterogeneity in FL imply different communication complexities and require technical novelties over results in distributed and decentralized optimization? These aspects are currently unexplored or underexplored in FL. We fill in the gaps here by providing affirmative answers.

\begin{table*}[th!]
\centering
\begin{tabular}{|c|c|c|c|c|}\hline 
	\multirow{2}{*}{\backslashbox{{\tiny Participation} }{{\tiny Objective function}}} & \multirow{2}{*}{Strongly Convex}     &\multirow{2}{*}{Convex}    & Overparameterized & Overparameterized \\ 
	                                &                        &         &     general case                 & linear regression   \\ \hline 
	Full                         & $\cO(\frac{1}{NT}+\frac{E^{2}}{T^{2}})$    &  $\mathcal{O}\left(\frac{1}{\sqrt{NT}}+\frac{NE^{2}}{T}\right)$   & $\cO(\exp(-\frac{NT}{E\kappa_1}))$ & $\cO(\exp(-\frac{NT}{E\kappa }))^{\dagger}$    \\ \hline
	Partial                      &  $\cO\left(\frac{E^{2}}{KT}+\frac{E^{2}}{T^{2}}\right)$   &  $\cO\left(\frac{E^2}{\sqrt{KT}}+\frac{KE^2}{T} \right)$ &  $\cO(\exp(-\frac{KT}{E\kappa_1}))$ & $\cO(\exp(-\frac{KT}{E\kappa }))^{\dagger}$     \\ \hline
\end{tabular}
\caption{Our convergence results for FedAvg and accelerated FedAvg in this paper. Throughout the paper, $N$ is the total
number of local devices, and $K \leq N$ is the maximal number of devices that are accessible to the central server. $T$ is the total number of stochastic updates performed by each local device, $E$ is the local steps between two consecutive server communications (and hence $T/E$ is the number of communications). $^{\dagger}$ In the linear regression setting, we have $\kappa=\kappa_1$ for FedAvg and $\kappa=\sqrt{\kappa_1\tilde{\kappa}}$ for momentum accelerated FedAvg (FedMaSS), where $\kappa_1$ and $\sqrt{\kappa_1\tilde{\kappa}}$ are condition numbers defined in Section \ref{sec:app:overparameterized}. Since
$\kappa_{1}\geq\tilde{\kappa}$, this implies a speedup factor of
$\sqrt{\frac{\kappa_{1}}{\tilde{\kappa}}}$ for accelerated FedAvg.}
% {\raggedright 
% $^{*}$ The fourth column presents the results for the general overparaterized setting. The fifth column presents the results for overparaterized linear regression.
        %   \par}
\label{tb:convergencerateintro}
\end{table*}

\textbf{Our Contributions}
We provide a comprehensive and unified convergence analysis
of FedAvg and its accelerated variants in the presence of both data and system heterogeneity. 
Our contributions are threefold.

First, we establish an {\small{$\cO(1/KT)$}} convergence rate  under FedAvg for strongly convex and smooth problems and  an
{\small{$\cO(1/\sqrt{KT})$}} convergence rate for convex
and smooth problems (where $K$ is the number of participating devices), thereby establishing that FedAvg enjoys the desirable linear speedup property in the FL setup. Prior to our work here, the best and the most related convergence analysis is given by \cite{li2019convergence} and~\cite{karimireddy2019scaffold}, which established an $\cO(\frac{1}{T})$ convergence rate for strongly convex smooth problems under FedAvg. Our rate matches the same (and optimal) dependence on $T$, but also completes the picture by establishing the linear dependence on $K$, for any $K\leq N$, where $N$ is the total number of devices, whereas~\cite{li2019convergence} does not have linear speedup analysis, and~\cite{karimireddy2019scaffold} only allows linear speedup close to full participation ($K=\mathcal{O}(N)$). As for convex and smooth problems, there was no prior work that established the {\small{$\cO(\frac{1}{\sqrt{T}})$}} rate under both system and data heterogeneity. Our unified analysis highlights the common elements and distinctions between the strongly and convex settings.

Second, we establish the same convergence rates--{\small{$\cO(1/KT)$}} for strongly convex and smooth problems and {\small{$\cO(1/\sqrt{KT})$}} for convex and smooth problems--for Nesterov accelerated FedAvg. 
We analyze the accelerated version of FedAvg here because empirically it tends to perform better; yet, its theoretical convergence guarantee is unknown. To the best of our knowledge, these are the first results that provide a linear speedup characterization of Nesterov accelerated FedAvg in those two problem classes (that FedAvg and Nesterov accelerated FedAvg share the same convergence rate is to be expected: this is the case even for centralized stochastic optimization). Prior to our results here, the most relevant results~\cite{yu2019linear,li2018federated,huo2020faster} only concern the non-convex setting, where convergence is measured with respect to stationary points (vanishing of gradient norms, rather than optimality gaps). Our unified analysis of Nesterov FedAvg also illustrates the technical similarities and distinctions compared to the original FedAvg algorithm, whereas prior works (in the non-convex setting) were scattered and used different notations.

Third, we study a subclass of strongly convex smooth problems where the objective is over-parameterized and establish 
a faster $\cO(\exp(-\frac{KT}{\kappa}))$ convergence rate for FedAvg, in contrast to the $\cO(\exp(-\frac{T}{\kappa}))$ rate for individual solvers~\cite{ma2017power}. Within this class, we further consider the linear regression problem and establish an even sharper rate under FedAvg. In addition, we propose a new variant of accelerated FedAvg based on a momentum update of~\cite{liu2018accelerating}--MaSS accelerated FedAvg--and establish a faster convergence rate (compared to if no acceleration is used). This stands in contrast to generic (strongly) convex stochastic problems where theoretically no rate improvement is obtained when one accelerates FedAvg.
The detailed convergence results are summarized in Table~\ref{tb:convergencerateintro}.

\textbf{Connections with Distributed and Decentralized Optimization}
Federated learning is closely related to distributed and decentralized optimization, and as such it is important to discuss connections and distinctions between our work and related results from that literature. First, when there is neither system heterogeneity, i.e. all devices participate in parameter averaging during a communication round, nor statistical heterogeneity, i.e. all devices have access to a common set of stochastic gradients, FedAvg coincides with the ``Local SGD'' of~\cite{stich2018local}, which showed the linear speedup rate $\mathcal{O}(1/NT)$ for strongly convex and smooth functions. \cite{woodworth2020local} and \cite{woodworth2020minibatch} further improved the communication complexity that guarantees the linear speedup rate. When there is only data heterogeneity, some works have continued to use the term Local SGD to refer to FedAvg, while others subsume it in more general frameworks that include decentralized model averaging based on a network topology or a mixing matrix. They have provided linear speedup analyses for strongly convex and convex problems, e.g.~\cite{khaled2020tighter,koloskova2020unified} as well as non-convex problems, e.g.~\cite{jiang2018linear,yu2019parallel,wang2018cooperative}. However, these results do not consider system heterogeneity, i.e. the presence of stragglers in the device network. Even with decentralized model averaging, the assumptions usually imply that model averages over all devices is the same as decentralized model averages based on network topology (e.g.~\cite{koloskova2020unified} Proposition 1), which precludes system heterogeneity as defined in this paper and prevalent in FL problems. For momentum accelerated FedAvg,~\cite{yu2019linear} provided linear speedup analysis for non-convex problems, while results for strongly convex and convex settings are entirely lacking, even without system heterogeneity.~\cite{karimireddy2019scaffold} considers both types of heterogeneities for FedAvg, but their rate implies a linear speedup only when the number of stragglers is negligible. In contrast, our linear speedup analyses consider both types of heterogeneity present in the full federated learning setting, and are valid for any number of participating devices. We also highlight a distinction in communication efficiency when system heterogeneity is present. Moreover, our results for Nesterov accelerated FedAvg completes the picture for strongly convex and convex problems. For a detailed comparison with related works, please refer to Table~\ref{tb:convergenceratev3} in Appendix Section~\ref{sec:app:comparison}.

\textbf{New related works since last time}: \cite{zhu2019multi,xu2020ternary,yang2020federated,chen2019communication,sattler2020clustered,konevcny2016federated,yuan2020stochastic,dinh2020federated,li2020acceleration,nedic2009distributed,duchi2011dual,liu2016collective,yuan2016convergence,lu2011gossip,ram2010distributed,yuan2018optimal,horvath2019stochastic,wangni2017gradient,alistarh2017qsgd}
